Bayesian inference of material culture phylogenies using continuous traits from the archaeological record: A birth–death model for Late Neolithic/Early Bronze Age arrowheads from Northwestern Europe

### 2023-07-05 analysis/paper/paper.qmd
remove data

We demonstrate the use of artefact shape variables as continuous characters for phylogenetic modelling using flaked stone arrowheads found in late Neolithic chiefly burials across north-west France, southern Britain and Denmark during 2500 BC and 1700 BC [4500--3700 cal BP, @nicolasArrowsPowerBrittany2017]. These arrowheads, sometimes found in wooden boxes or wrapped in leather, are part of suite of grave goods that include bell-shaped ceramic food vessels, ornaments, and tools made from amber, bronze, copper, gold, silver and other materials. This set of grave goods is known as the Bell Beaker culture that spread across Europe through a mixture of migration and cultural diffusion. The arrowheads were bifacially flaked using antler, bone, and metal indentors, and many have traces of adhesive residue used for halfting onto a shaft. The usewear traces and geometry of halfting indicate that the arrowheads were loosely attached, designed for symbolic displace rather than efficient shots [@nicolasArrowsPowerBrittany2017]. Diagnostic impact fractures are generally absent from the arrowheads, with less than three percent showing signs of having been shot, indicating that most were not used as projectiles, but as non-functional display items. They were likely made by specialist master craft workers, with the production sequence transmitted through generations, and the final products soley intended to portray and emphasise the social status of the individuals they were buried with. Given that the arrowheads were primarily symbols that were not used as projectiles, we can be confident that shape variation is unrelated to physical performance factors, and instead likely reflects transmission of social values and identities.

```{r}
#| label: fig-pca-arrowheads
#| fig-cap: "The first five principal components from our PCA showing how they capture distinct shape . PC1: tang depth, PC2: shoulder width, PC3: tang height, PC4: tang width, PC5: shoulder height "

@@

We focus on the `r n_artefacts_france_with_chrono_info` arrowheads from sites in France because these have the most complete provenance and chronological information. @nicolasFlechesPouvoirAube2016 provided detailed images of the arrowheads, and a techno-typological classification of these arrowheads projected on the period\'s chronology, based on metric measurements and morphological observations. We started with the outlines extracted from these images by @matzig2021 and followed their procedure to remove outliers, and compute an EFA to quantitatively describe the outlines as a set of harmonics. We used the `r n_harmonics_for_99.9` harmonics required to capture 99.9% of shape variation. To improve computational efficiency, we reduced the dimensionality of these these harmonics by inputing them into a PCA, keeping the `r pc_for_99` components need to capture 99% of variation in the harmonics. These components are our shape variables that we use as continuous traits to input into our phylogenetic models. @fig-pca-arrowheads illustrates how the principal components capture distinct shape characteristics.

We generated fours models to investigate the most appropriate combination of substitution, clock, and tree models. @tbl-model-summary summarises the model configurations and their marginal likelihoods. Taking the first model as the null model, the Bayes Factors indicate strong support for the model using a birth-death tree (where variants can appear and disappear, rather than only appear), relaxed Brownian motion (where the rate of morphological change is allowed to vary among branches of the phylogeny) and a relaxed clock (allowing for shifts in rates of morphological change across branches).

@@

knitr::kable(model_summary_tbl_nums,
             caption = "Summary of phygenetic models explored with the arrowhead data")
```

## Conclusion

>>>>

We demonstrate the use of artefact shape variables as continuous characters for phylogenetic modelling using flaked stone arrowheads found in late Neolithic chiefly burials across north-west France, southern Britain and Denmark during 2500 BC and 1700 BC [4500--3700 cal BP, @nicolasArrowsPowerBrittany2017]. These arrowheads, sometimes found in wooden boxes or wrapped in leather, are part of suite of grave goods that include bell-shaped ceramic food vessels, ornaments, and tools made from amber, bronze, copper, gold, silver and other materials. This set of grave goods is known as the Bell Beaker culture that spread across Europe through a mixture of migration and cultural diffusion. The arrowheads were bifacially flaked using antler, bone, and metal indentors, and many have traces of adhesive residue used for halfting onto a shaft. The usewear traces and geometry of halfting indicate that the arrowheads were loosely attached, designed for symbolic displace rather than efficient shots [@nicolasArrowsPowerBrittany2017]. Diagnostic impact fractures are generally absent from the arrowheads, with less than three percent showing signs of having been shot, indicating that most were not used as projectiles, but as non-functional display items. They were likely made by specialist master craft-workers, with the production sequence transmitted through generations, and the final products soley intended to portray and emphasise the social status of the individuals they were buried with. Given that the arrowheads were primarily symbols that were not used as projectiles, we can be confident that shape variation is unrelated to physical performance factors, and instead likely reflects transmission of social values and identities.

```{r}
#| label: fig-pca-arrowheads
#| fig-cap: "The first five principal components from our PCA showing how they capture distinct shape. PC1: tang depth, PC2: shoulder width, PC3: tang height, PC4: tang width, PC5: shoulder symmetry"

@@

We focus on the `r n_artefacts_france_with_chrono_info` arrowheads from sites in France because these have the most complete provenance and chronological information. @nicolasFlechesPouvoirAube2016 provided detailed images of the arrowheads, and a techno-typological classification of these arrowheads projected on the period's chronology, based on metric measurements and morphological observations. We started with the outlines extracted from these images by @matzig2021 and followed their procedure to remove outliers, and compute an EFA to quantitatively describe the outlines as a set of harmonics. We used the `r n_harmonics_for_99.9` harmonics required to capture 99.9% of shape variation. To improve computational efficiency, we reduced the dimensionality of these these harmonics by inputing them into a PCA, keeping the `r pc_for_99` components need to capture 99% of variation in the harmonics. These components are our shape variables that we use as continuous traits to input into our phylogenetic models. @fig-pca-arrowheads illustrates how the principal components capture distinct shape characteristics.

We generated fours models to investigate the most appropriate combination of substitution, clock, and tree models. @tbl-model-summary summarises the model configurations and their marginal likelihoods. Full details of our Bayesian MCMC procedure are documented in our supplementary materials at http://doi.org/xxx/xxx. Taking the first model as the null model, the Bayes Factors indicate strong support for the model using a birth-death tree (where variants can appear and disappear, rather than only appear), relaxed Brownian motion (where the rate of morphological change is allowed to vary among branches of the phylogeny) and a relaxed clock (allowing for shifts in rates of morphological change across branches). @fig-treemap-densitree-map summarises the phylogenetic landscape, with the top left panel showing a single cluster of related trees, indicating that one phylogeny is definitively supported by the data [@jombartTreespaceStatisticalExploration2017]. The top right panel is a Densitree plot of trees randomly sampled from the converged posterior to provide intuition about the extent of incongruence among these trees [@bouckaertDensiTreeMakingSense2010]. It shows high congruence for the Bronze A2 and Bell Beaker samples, but substantial uncertainty for the Bronze A1 arrowheads, indicating competing topologies and possibly horizontal transmission and independent convergence on similar shapes.

@@

knitr::kable(model_summary_tbl_nums)
```

The middle and lower panel of @fig-treemap-densitree-map demonstrates the presence of a phylogenetic signal in our data. This is the tendency for related artefacts to resemble each other more than they resemble species drawn at random from the tree [@blombergTestingPhylogeneticSignal2003]. The middle panel shows the maximum a posteriori topology which is the tree that has the greatest posterior probability, averaged over all branch lengths and substitution parameter values. The middle panel also shows the strength of phylogenetic signal for each shape variable via the Local Indicator of Phylogenetic Association (local Moran's I) for each specimen and trait to indicate hotspots of association [@keckPhylosignalPackageMeasure2016]. The bottom panel of @fig-treemap-densitree-map show's Pagel's lambda values for our traits, nearly all are \~1 with p \< 0.05, indicating that there is a significant phylogenetic signal for all traits in our model, compared to a random model of evolution (lambda \~ 0).

```{r}
#| label: fig-treemap-densitree-map
#| fig-cap: "Exploration of landscapes of phylogenetic trees. Top left: Pairwise Euclidean distances between trees decomposed into a low-dimensional space using multidimensional scaling (MDS); top right: Densitree of all trees in the set drawn transparently, areas where a lot of the trees agree in topology and branch lengths show up as highly colored areas, while areas with little agreement show up as webs; middle panel:local Moran's I for each shape variable to detect hotspots of autocorrelation, red bars indicate significant I values (p < 0.01), indicating phylogenetic signal.; bottom panel: Pagel's lamba values for each shape variable, all have p < 0.05."
#| fig-height: 10
#| fig-width: 7
#| out-height: "500"

knitr::include_graphics(here::here("analysis/figures/treespace-densitree-map-signal-lamdba.png"))

```

Overall the tree is consistent with the traditional scheme of typological evolution presented by @nicolasFlechesPouvoirAube2016 in showing the Bell Beaker and three major phases of Bronze Age production as mostly coherent clades. Arrowheads from Early Bronze Age 2 and 3 show much greater coherence than those from Early Bronze Age 1, with several find locations appearing in a single clade. One interpretation of this could be that during EBA-1, arrowheads for a single burial were sourced from a variety of craft workers, representing different lineages of arrow making. Later, in EBA-2 and 3, most or all of the arrowheads placed in a burial were sourced from a single master craft worker.

The relaxed clock model provides insights into how rates of change in artefact shape varied across the tree. The left panel of @fig-tree-rates-morpho shows that rates of evolution are mostly fairly even across most the tree, with the slowest rates in an EBA-1 clade and the fastest rates in an EBA-2 clade. One interpretation of this is that the craft workers that made the earliest arrowheads were conservative in their exploration of shape possibilities as they were making their first appearance, then briefly experimented with a faster rate of change as use of the arrowheads became more established. Perhaps a conformist bias was at work as the arrowheads initially became established as burial goods. The right panel of @fig-tree-rates-morpho shows the distribution of trait values across the tree for the first five principal components, as our primary shape variables, and projections of the tree onto the morphospace for each combination of these variables. This shows that the deep tang (PC1) and narrow shoulders (PC2) were characteristics typical of the clades representing the middle of study period. We can also see that the area of the tree with highest rates of change has extreme values of (PC1: deep tang), indicating that much of the change occurred in that trait.

```{r}
#| label: fig-tree-rates-morpho
#| fig-cap: "Left: Rates of evolution across the tree; Right: multidimensional phylogenetic phylomorphospace scatterplot matrix"
#| fig-height: 10
#| fig-width: 7
#| out-height: "500"

knitr::include_graphics(here::here("analysis/figures/tree-rates-morphospace.png"))

```

## Discussion and Conclusion

Our brief case study demonstrates an accessible workflow using open source software tools for reconstructing phylogenetic relationships among archaeological artefacts. We have demonstrated how continuous characters, in the form of shape variables derived from geometric morphometry, which have become a popular way of quantitatively describing complex artefacts [@matzig2023], can be input into a phylogenetic model to yield easily interpretable output. The arrowheads of our case study are suitable because their shape did not change over the use-life of the artefact -- they were produced as grave goods and not used as projectiles. Other artefacts made using reductive techniques, such as Mousterian scrapers, might not be suitable for this kind of analysis because their shape changes substantially over their use-life [@dibbleInterpretationMiddlePaleolithic1987], independent of vertical transmission of cultural information about artefact shape.

Our use of Bayesian methods has provided formal approaches to manage competing models about the processes that generated the data, and methods for formalising uncertainty. These can be contrasted with previous work using parsimony methods that typically do not convey error estimates. However, Bayesian methods presents new challenges in the selection of appropriate models and prior parameter distributions. We anticipate that ethnographic and detailed archaeological data on transmission dynamics may help here.

This brief case study yielded promising results and sketches a straightforward template for future work in archaeological phylogenetics. We have demonstrated a model-based Bayesian workflow, and how to convey uncertainty, obtain insights about phylogenetic signal, and rates and locations of change. This case study could be extended with samples from a larger area, for example South Britain and Denmark @nicolasFlechesPouvoirAube2016], and by parallel analyses of other artefact classes found with the burials. More generally whole 'society' transmission dynamics or contrasting transmission across genders could also be explored with these methods. At much broader scales, phylogenetic/macroevolutionary studies of gene-culture co-evolution may be possible, as well as exploration of cultural phylogeography and niche construction dynamics using comparative methods.
